
###Reproducible Research : Peer Assessment 1

```{r libraries}
library(ggplot2) 
library(scales)
library(Hmisc)
```


### Load and preprocess the data 
#### 1. Load the data (read.csv())
```{r load}
if (!file.exists('activity.csv')) {
  unzip('activity.zip')
}
activityData <- read.csv('activity.csv')

```
#### 2.Process the raw data 
```{r process}
activityData$date <- as.Date(activityData$date, format = "%Y-%m-%d")
activityData$interval <- as.factor(activityData$interval)
```
#### 3.Check the Data
```{r check}
str(activityData)
```
### What is the mean Total number of steps taken per day?
```{r histogram}
###calculate the steps per day
steps_per_day <- aggregate(steps ~ date, activityData, sum)
colnames(steps_per_day) <- c("date","steps")
head(steps_per_day)

###Histogram of the total number of steps taken each day
ggplot(steps_per_day, aes(x = steps)) + 
       geom_histogram(fill = "blue", binwidth   = 1000) + 
        labs(title="Histogram of the Total Steps Taken Each Day", 
             x = "Number of Steps Each Day", y = "Number of times in a day(Count)") + theme_bw() 
```
```{r mean_median }
### Mean and median number of steps taken each day
steps_mean   <- mean(steps_per_day$steps, na.rm=TRUE)
steps_median <- median(steps_per_day$steps, na.rm=TRUE)

```

#### The Mean is : `r steps_mean` and Median is : `r steps_median`

Time series plot of the average number of steps taken
```{r  time}
###calculate the aggregation of steps by intervals of 5-minutes
steps_per_interval <- aggregate(activityData$steps, 
                                by = list(interval = activityData$interval),
                                FUN=mean, na.rm=TRUE)
###convert to integers
steps_per_interval$interval <- 
        as.integer(levels(steps_per_interval$interval)[steps_per_interval$interval])
colnames(steps_per_interval) <- c("interval", "steps")

###Plot with the time series of the average numbers of taken across all the days 
ggplot(steps_per_interval, aes(x=interval, y=steps)) +   
        geom_line(color="blue", size=1) +  
        labs(title="Average Daily Activity Pattern", x="5 minute Interval", y="Number of steps") +  
        theme_bw()
```

The 5-minute interval that, on average, contains the maximum number of steps
```{r minute}
max_interval <- steps_per_interval[which.max(steps_per_interval$steps),]
colnames(max_interval) <- c("interval","steps")
```
#### The `r max_interval$interval` interval has maximum  `r max_interval$steps` steps 

Imputing missing values
 1. The total number of missing values in the dataset
```{r}
missing_values <- sum(is.na(activityData$steps))
```
The total number of missing values are `r missing_values`

2. strategy for filling in all of the missing values in the dataset.
3. Create new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
activity_data_imputed <- activityData
activity_data_imputed$steps <- impute(activityData$steps, fun=mean)
```
#### 4. Plot histogram of the total number of steps taken each day
```{r}
steps_by_day_imputed <- tapply(activity_data_imputed$steps, activity_data_imputed$date, sum)
qplot(steps_by_day_imputed, xlab='Total steps per day ', ylab='Frequency using binwith 500', binwidth=500)

```
#### Calculate and report the mean and median total number of steps taken per day
```{r}
steps_by_mean_imputed <- mean(steps_by_day_imputed)
steps_by_median_imputed <- median(steps_by_day_imputed)
```
#### Mean : `r steps_by_mean_imputed` and Median : `r steps_by_median_imputed`

Are there differences in activity patterns between weekdays and weekends?
1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r}
activity_data_imputed$wtype <-  ifelse(as.POSIXlt(activity_data_imputed$date)$wday %in% c(0,6), 'weekend', 'weekday')

```
#### Make a time series plot
```{r}
average_activity_data_imputed <- aggregate(steps ~ interval + wtype, data=activity_data_imputed, mean)
ggplot(average_activity_data_imputed, aes(x=interval, y=steps)) + 
    geom_point() + 
    facet_wrap(~ wtype ) +
    xlab("5-minute interval") + 
    ylab("avarage number of steps") +
    theme_bw()
```
